High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-08-29 paper/main.tex
working on methods section, up to fmri section

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as 125 participants listened to either
an audio recording of a ten-minute story or a temporally scrambled
version of the story, or underwent a resting state
scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of correlations between those patterns that were
inferred using our computational framework).  We found that both the
PCA-based and eigenvector centrality-based approaches yielded neural
patterns that could be used to decode accurately.  Both approaches
also yielded the best decoding accuracy for data collected during
(intact) story listening when high-order (PCA:
second-order; eigenvector centrality: fourth-order) dynamic
correlation patterns were included as features.  When we trained
classifiers on the scambled stories or resting state data, only
lower-order dynamic patterns were informative to the decoders.  Taken
together, our results indicate that high-level cognition is supported
by high-order dynamic patterns of communication between brain
structures.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/methods_fig}
  \caption{\textbf{Estimating dynamnic high-order correlations.}  Given
    a $T$ by $K$ matrix of multivariate timeseries data,
    $\mathbf{Y}_n$ (where $n \in \mathbb{N}, n \geq 0$), we use
    Equation~\ref{eqn:timecorr} to compute a timeseries of
    $K$ by $K$ correlation matrices, $\mathbf{Y}_{n+1}$.  We then
    approximate $\mathbf{Y}_{n+1}$ with the $T$ by $K$ matrix
    $\mathbf{X}_{n+1}$.  This process may be repeated to scalably estimate
    iteratively higher-order correlations in the data.}
  \label{fig:methods}
\end{figure}

@@

For a given graph theoretic measure,
$\eta: \mathcal{R}^{K \times K} \rightarrow \mathbb{R}^K$, we can use
$\eta$ to tranform each row of $\mathbf{Y}_n$ in a way that
characterizes the corresponding graph-theoretic properties of each
column.  This results in a new $T$ by $K$ matrix, $\mathbf{X}_n$, that
reflects how the features reflected in the columns of $\mathbf{Y}_n$
participate in the network during each timepoint (row).

@@

Because $\mathbf{X}_n$ has the same shape as the original data
$\mathbf{X}_0$, approximating $\mathbf{Y}_n$ with a lower-dimensional
$\mathbf{X}_n$ enables us to estimate high-order dynamic correlations
in a scalable way.  Given a $T$ by $K$ input matrix, the output of
Equation~\ref{eqn:timecorr} requires $\mathcal{O}(TK^2)$ space to
store.  Repeated applications of Equation~\ref{eqn:timecorr} (i.e.,
computing dynamic correlations between the columns of the outputted
dynamic correlation matrix) each require exponentially more space; in
general the $n^\mathrm{th}$-order dynamic correlations of a $T$ by $K$
timeseries occupies $\mathcal{O}(TK^{2n})$ space.  However, when we
approximate or summarize the output of Equation~\ref{eqn:timecorr} with a $T$ by
$K$ matrix (as described above), it becomes feasible to compute even
very high-order correlations in high-dimensional data.  Specifically,
approximating the $n^\mathrm{th}$-order dynamic correlations of a $T$
by $K$ timeseries requires only $\mathcal{O}(TK^2)$ additional space--
the same as would be required to compute first-order dynamic
correlations. In other words, the space required to store $n+1$
multivariate timeseries reflecting up to $n^\mathrm{th}$ order
correlations in the original data scales linearly with $n$ using our
approach (Fig.~\ref{fig:methods}). \textbf{JRM STOPPED HERE}

@@

\subsubsection*{Synthetic data}
\paragraph*{Constant.}
\paragraph*{Random.}
\paragraph*{Ramping.}
\paragraph*{Event.}

>>>>

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as participants
listened to either an audio recording of a ten-minute story or a
temporally scrambled version of the story, or underwent a resting
state scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of correlations between those patterns that were
inferred using our computational framework).  We found that both the
PCA-based and eigenvector centrality-based approaches yielded neural
patterns that could be used to decode accurately.  Both approaches
also yielded the best decoding accuracy for data collected during
(intact) story listening when high-order (PCA: second-order;
eigenvector centrality: fourth-order) dynamic correlation patterns
were included as features.  When we trained classifiers on the
scambled stories or resting state data, only lower-order dynamic
patterns were informative to the decoders.  Taken together, our
results indicate that high-level cognition is supported by high-order
dynamic patterns of communication between brain structures.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/methods_fig}
  \caption{\textbf{Estimating dynamnic high-order correlations.}  Given
    a $T$ by $K$ matrix of multivariate timeseries data,
    $\mathbf{Y}_n$ (where $n \in \mathbb{N}, n \geq 0$), we use
    Equation~\ref{eqn:timecorr} to compute a timeseries of
    $K$ by $K$ correlation matrices, $\mathbf{Y}_{n+1}$.  We then
    approximate $\mathbf{Y}_{n+1}$ with the $T$ by $K$ matrix
    $\mathbf{X}_{n+1}$.  This process may be repeated to scalably estimate
    iteratively higher-order correlations in the data.  Note that the
    transposes of $\mathbf{Y}_n$ and $\mathbf{X}_{n+1}$ are displayed
    in the figure for compactness.
  \label{fig:methods}}
\end{figure}

@@

For a given graph theoretic measure,
$\eta: \mathbb{R}^{K \times K} \rightarrow \mathbb{R}^K$, we can use
$\eta$ to tranform each row of $\mathbf{Y}_n$ in a way that
characterizes the corresponding graph-theoretic properties of each
column.  This results in a new $T$ by $K$ matrix, $\mathbf{X}_n$, that
reflects how the features reflected in the columns of $\mathbf{Y}_n$
participate in the network during each timepoint (row).

@@

Because $\mathbf{X}_n$ has the same shape as the original data
$\mathbf{X}_0$, approximating $\mathbf{Y}_n$ with a lower-dimensional
$\mathbf{X}_n$ enables us to estimate high-order dynamic correlations
in a scalable way.  Given a $T$ by $K$ input matrix, the output of
Equation~\ref{eqn:timecorr} requires $\mathcal{O}(TK^2)$ space to
store.  Repeated applications of Equation~\ref{eqn:timecorr} (i.e.,
computing dynamic correlations between the columns of the outputted
dynamic correlation matrix) each require exponentially more space; in
general the $n^\mathrm{th}$-order dynamic correlations of a $T$ by $K$
timeseries occupies $\mathcal{O}(TK^{2n})$ space.  However, when we
approximate or summarize the output of Equation~\ref{eqn:timecorr} with a $T$ by
$K$ matrix (as described above), it becomes feasible to compute even
very high-order correlations in high-dimensional data.  Specifically,
approximating the $n^\mathrm{th}$-order dynamic correlations of a $T$
by $K$ timeseries requires only $\mathcal{O}(TK^2)$ additional space--
the same as would be required to compute first-order dynamic
correlations. In other words, the space required to store $n+1$
multivariate timeseries reflecting up to $n^\mathrm{th}$ order
correlations in the original data scales linearly with $n$ using our
approach (Fig.~\ref{fig:methods}).

@@

\subsubsection*{Synthetic data}
We constructed a total of 40 multivariate timeseries, collectively
reflecting a total of 4 different patterns of dynamic correlations
(i.e., 10 datasets reflecting each type of dynamic pattern).  Each
timeseries comprised 50 features (dimensions) that varied over 300
timepoints.  The observations at each timepoint were drawn from a
zero-mean multivariate Gaussian distribution with a covariance matrix
defined for each timepoint as described below.  We drew the observations at
each timepoint independently from the draws at all other timepoints;
in other words, for each observation $s_t \sim
\mathcal{N}\left(\mathbf{0}, \mathbf{\Sigma}_t\right)$ at timepoint $t$, $p(s_t) = p(s_t | p_{\backslash t})$.

\paragraph*{Constant.}  We generated data with stable underlying
correlations to evaluate how Equation~\ref{eqn:timecorr} characterized
correlation ``dynamics'' when the ground truth correlations were
static.  We constructed 10 multivariate timeseries, whose observations
were each drawn from a single (stable) Gaussian distribution.  For
each dataset, we constructed a random covariance matrix,
$\mathbf{\Sigma}_m$:
\begin{align}
  \mathbf{C}(i, j) &\sim \mathcal{N}(0, 1)\\
  \mathbf{\Sigma}_m &= \mathbf{C}
                                          \mathbf{C}^\top~\mathrm{, where}\label{eqn:randcorr}
  \end{align}
  $i, j \in \left[1, 2, ..., 50 \right]$.  In other words, all of the
  observations (for each of the 300 timepoints) within each dataset
  were drawn from a multivariate Gaussian distribution with the same
  covariance matrix, and the 10 datasets each used a different
  covariance matrix.

  \paragraph*{Random.}  We generated a second set of 10 synthetic
  datasets whose observations at each timepoint were drawn from a
  Gaussian distribution with a new randomly constructed (using
  Eqn.~\ref{eqn:randcorr}) covariance matrix.  Because each
  timepoint's covariance matrix was drawn independently of the
  covariance matrices for all other timepoints, these datasets
  provided a test of reconstruction accuracy in the absence of any
  meaningful underlying temporal structure in the dynamic correlations
  underlying the data.

  \paragraph*{Ramping.}  We generated a third set of 10 synthetic
  datasets whose underlying correlations changed gradually over time.
  For each dataset, we constructed two \textit{anchor} correlation
  matrices using Equation~\ref{eqn:randcorr},
  $\mathbf{\Sigma}_{\mathrm{start}}$ and
  $\mathbf{\Sigma}_{\mathrm{end}}$.  For each of the 300 timepoints in
  each dataset, we drew the observations from a multivariate Gaussian
  distribution whose covariance matrix at each timepoint $t \in
  \left[0, ..., 299\right]$ was given by
  \begin{align}
    \mathbf{\Sigma}_t = \left( 1 - \frac{1 - t}{299} \right)
    \mathbf{\Sigma}_{\mathrm{start}} + \frac{t}{299}\mathbf{\Sigma}_{\mathrm{end}}.
  \end{align}
The gradually changing correlations underlying these datasets allow us
to evaluate the recovery of dynamic correlations when each timepoint's
correlation matrix is unique (as in the random datasets), but where
the correlation dynamics are structured.

\paragraph*{Event.} We generated a fourth set of 10 synthetic datasets
whose underlying correlation matrices exhibited prolonged intervals of
stability, intersperced with abrupt changes.  For each dataset, we
used Equation~\ref{fig:randcorr} to generate 5 random covariance
matrices.  We constructed a timeseries where each set of 60 consecutive samples
was drawn from a Gaussian with the same covariance matrix.  These
datasets were intended to simulate a system that undergoes occasional
abrupt state changes.

\textbf{JRM STOPPED HERE}
